A Hybrid Neutronics Method for Time-Dependent Control Rod Modeling in Molten Salt Reactors

### 2024-06-24 hybrid.tex
final edits for first draft of chapter 4; add image files for chapter 2,3,4

Transport corrections to allow neutron diffusion methods to reproduce high-fidelity flux solutions
can be introduced by modifying the neutron diffusion equations.
In this work, I explored two options for applying
transport corrections to the neutron diffusion equations: applying
diffusion corrections or adding a drift correction term.
I performed preliminary investigations on 1-D test cases with both types of transport correction. I
eventually elected to use the drift correction term for the remaining test cases and reactor model
demonstrations due to superior numerical properties. I will make clear the reasons for this
selection through 1-D test case results in Chapter 5.

@@

This boundary condition for the $S_N$ subsolver will generally cause deviations
in the flux and transport correction values since the $P_1$ approximation is accurate for up to
linear anisotropies in the flux. However, as I will show in the next chapter, the influence of
boundary conditions on the accuracy of
the transport correction terms does not extend far from the boundary $\partial V_1$ in optically
thick media due to competing influences from the bulk scattering and absorption effects in $V_1$.
Results in the next chapter will show that transport corrections generated with the $S_N$
subsolver within $V_1$ are accurate everywhere up to a few centimeters near $\partial V_1$.

When applying the hybrid method, $V_1$ must be large enough to provide accurate transport
corrections near the highly-absorbing control rod while simultaneously accommodating for inaccurate
transport corrections near $\partial V_1$. I will henceforth refer to the region in $V_1$
near $\partial V_1$ with inaccurate transport corrections as the \textit{buffer region} ($V_1'$).
We will discard inaccurate transport correction values in $V_1'$. Therefore, the subsequent neutron
diffusion calculation on $V_0$ will only apply transport corrections in $V_1$ excluding $V_1'$.

@@

All source code for the hybrid $S_N$-diffusion model can be found in my Moltres GitHub fork
repository\footnote{\url{https://github.com/smpark7/moltres/tree/sn-method}} as of the time of
writing. This branch may be merged into the original upstream Moltres GitHub repository in future.

@@

The \gls{SAAF} $S_N$ method defines $G$ energy groups and $N_d$ discrete directions for a total
of $G\times N_d$ angular flux variables $\Psi_{g,d}$. The Moltres $S_N$ model groups the
$\Psi_{g,d}$ variables by energy group into $G$ \textit{array variables}. An array variable is a
set of standard field variables to help with dealing with setting up simulations with high
dimensionality. Each standard variable of an array variable is referred to as a component in the
array variable. In this work, I opted for
2nd-order Lagrange shape functions to approximate the $\Psi_{g,d}$ variables in the finite-element
analysis. The 2nd-order approximation showed better convergence rates in mesh refinement studies.

@@

The \gls{MOOSE} framework leverages the Eigen \cite{guennebaud_eigen_2010} C++ template library for
linear algebra operations defined in ArrayKernels and ArrayBCs for array variables. The \gls{SAAF}
$S_N$ kernels and BCs contain functions for computing the residual and Jacobian contributions of
each term. The \gls{FEM} numerical solver deems to have reached a converged solution if the sum of
residual contributions falls below a defined tolerance value. All $S_N$ simulations in this work
employ the \gls{PJFNK} method with the HYPRE-BoomerAMG method for preconditioning. BoomerAMG is an
algebraic multigrid method from the HYPRE \cite{hypre_hypre_2022} linear solver library.

The drift correction terms $\vec{D}_g$ are initialized as auxiliary array variables
with three components corresponding to their $x$, $y$, and $z$ components. Auxiliary variables
encompass all field variables which are not directly determined during the \gls{PJFNK}
calculation stage. The drift and boundary correction variables are computed during the
post-calculation stage from $\Psi_{g,d}$.

I employed the \texttt{MultiApp} system \cite{gaston_physics-based_2015} to implement the fixed point
iterative coupling between the $S_N$ and neutron diffusion solvers for the hybrid $S_N$-diffusion method.
The \texttt{MultiApp} and \texttt{Transfers} systems from \gls{MOOSE} provide flexible interfaces
for setting up iterative solvers and transferring data between the solvers. A typical hybrid
method simulation on Moltres starts with the neutron diffusion solver for the first estimate of the
neutron flux distribution. Moltres transfers the neutron flux distribution to the $S_N$ solver to
compute the boundary source distribution. The $S_N$ solver then generates its own flux estimate and
computes drift and boundary correction variables for the neutron diffusion solver. Moltres
transfers the correction variables to the neutron diffusion solver to compute its next flux
estimate. This process continues until the neutron flux distribution converges to within tolerance
values provided in the input file.

@@

In this chapter, I presented the mathematical derivations for the \gls{SAAF} formulation of the
$S_N$ method, transport correction formulations, and boundary conditions relevant to the hybrid
$S_N$-diffusion method. I also presented details on the iteration algorithm and the handling of
transport corrections in the hybrid $S_N$-diffusion method to improve on the standard neutron
diffusion method for neutronics simulations of reactor systems containing highly neutron-absorbing
regions (control rods). The hybrid method relies on transport corrections derived from the flux
solutions from the $S_N$ subsolver to fix inaccurate flux estimates from using the neutron
diffusion method alone. The hybrid method saves on computational expense at the cost of some
accuracy in reactor regions not treated with transport corrections. In the next chapter, I will
also show that I can further reduce computational expense by leaving the flux in the $S_N$
subsolver because the transport correction parameters converge faster than the flux in the $S_N$
subsolver.

%I established a framework for the hybrid method for coupling an $S_N$ subsolver and a neutron
%diffusion solver through outer iterations by taking advantage of the weak dependence of the $S_N$
%flux solution on the approximate boundary conditions determined from the neutron diffusion flux
%solution. This framework could be extended to improve multischeme methods, which
%divide a problem domain into subdomains where different neutronics methods are applied. For
%instance, in an $S_N$-diffusion multischeme method, the $S_N$ method is applied in subdomains with
%significant transport effects while other subdomains are treated with the neutron diffusion method
%to reduce the overall computational effort required \cite{wang_rattlesnake_2021}. 
%
%I also developed an algorithm for identifying and discarding inaccurate \gls{SVDC} values, which
%are expected to occur near the correction region boundary. I demonstrated the hybrid
%method on 1-D graphite-moderated test cases modeled after the \gls{MSRE}. The hybrid method showed
%good agreement with the reference Monte Carlo and $S_8$ methods after eliminating neutron energy
%group discretization errors in test cases that included highly neutron-absorbing regions. The
%hybrid method had better accuracy in more homogeneous systems largely due to the larger effective
%transport correction regions through \glspl{SVDC}. Further work
%is needed to extend and verify the hybrid method to 2-D and 3-D neutronics simulations and to
%create a robust solution for determining \glspl{SVDC} around neutron flux peaks and troughs.

>>>>

Transport corrections to allow neutron diffusion methods to reproduce high-fidelity flux solutions
can be introduced by modifying the neutron diffusion equations.
In this work, I explored two options for applying
transport corrections to the neutron diffusion equations: applying
diffusion corrections or adding a drift correction term.
I performed preliminary investigations on 1-D test cases with both types of transport correction. I
eventually elected to use the drift correction term for the remaining test cases and reactor model
demonstrations due to superior numerical properties. I will make clear the reasons for this
selection through 1-D test case results in Chapter \ref{chap:hybrid-verification}.

@@

The $S_N$ subsolver boundary conditions will generally cause deviations
in the flux and transport correction values since the $P_1$ approximation is accurate for up to
linear anisotropies in the flux. However, as I will show in Chapter \ref{chap:hybrid-verification},
the influence of boundary conditions on the accuracy of
the transport correction terms does not extend far from the boundary $\partial V_1$ in optically
thick media due to competing influences from the bulk scattering and absorption effects in $V_1$.
Results in Chapter \ref{chap:hybrid-verification} will show that transport correction values
($D^s_g$ or $\vec{D}_g$) generated with the $S_N$ subsolver within $V_1$ are accurate everywhere up
to a few centimeters away from $\partial V_1$.

For accurate flux estimates with the hybrid method, I propose that $V_1$ must be large enough to
provide accurate transport corrections near the highly-absorbing control rod while simultaneously
accommodating for inaccurate transport correction values near $\partial V_1$. Right before each
neutron diffusion calculation, the solver must determine where the inaccurate transport correction
values occur and discard them. I will refer to the region near $\partial V_1$ in which the
transport correction values are discarded as the \textit{buffer region} ($V_1'$).
$V_1'$ shares the same outer boundary $\partial V_1$ as $V_1$. Consequently, the neutron
diffusion calculation on $V_0$ will only apply transport corrections in $V_1$ excluding $V_1'$.

In this discussion thus far, I have yet to establish the cutoff boundary for $V_1'$ within $V_1$
opposite to $\partial V_1$. A natural/intuitive criterion for the location of the cutoff boundary
would be wherever the components of the drift correction variable $\vec{D}_g$ is zero, i.e.,
wherever the components change signs. The reasons for this criterion are twofold. Firstly, at
points where the $\vec{D}_g$ components are zero, the flux is approximately isotropic along the
axes corresponding to the components because value of $\vec{D}_g$ represents how much correction a
standard neutron diffusion model requires. Secondly, this choice preserves the smoothness of the
neutron flux and flux gradient.
%
\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\columnwidth]{case-3b-geometry}
	\caption{1-D geometry for Case 3b.}
	\label{fig:3b-geometry}
\end{figure}

Here, I use Case 3b of the 1-D test cases to be covered in Chapter \ref{chap:hybrid-verification}
to illustrate the buffer region and the behavior of $\vec{D}_g$. Case 3b is a 1-D slab problem
consisting of control rod, air, fuel-graphite repeating lattice, and core vessel regions as shown
in Figure \ref{fig:3b-geometry}. The fuel regions are neutron-multiplying media containing
$^\text{235}$U fissile material. I took material compositions of each region from the \gls{MSRE}
reference design and the start-up fuel composition \cite{fratoni_molten_2020}.
Case 3b has reflecting boundary conditions on the left boundary and vacuum boundary conditions on
the right boundary.
%
\begin{figure}[htb!]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{case-3b-group-1-drift}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{case-3b-group-8-drift}
    \end{subfigure}
    \caption{The reference and hybrid drift ($\vec{D}_g$) distributions for group 1 and 8 calculated
      from $S_8$ and $S_8$-diffusion simulations. The correction region $V_1$ spans $x=0$ cm to
      $x=10$ cm.}
    \label{fig:3b-drift}
\end{figure}

I ran an eight energy group $S_8$ $k$-eigenvalue simulation with the newly implemented $S_N$ solver
on Moltres to generate reference $\vec{D}_g$ distributions. I also ran an eight-group hybrid
$S_8$-diffusion simulation with the $S_8$ method for generating $\vec{D}_g$ in $V_1$. I set $V_1$
to encompass the control rod region and its vicinity from $x=0$ to $10$ cm based on the geometry in
Figure \ref{fig:3b-geometry}. Figure \ref{fig:3b-drift} shows the $\vec{D}_1$ and $\vec{D}_8$
distributions from $x=0$ to $15$ from the reference $S_8$ and hybrid simulations. The hybrid
$\vec{D}_g$ distribution is zero beyond $x=10$ cm because no corrections are generated outside of
$V_1$. Comparing the hybrid distributions to the reference distributions in both energy groups, the
hybrid distributions are accurate up to around $x=8$ cm. Working backward towards the control rod
at $x=0$-$0.5$ cm, the $\vec{D}_g$ values go to zero at $x=7.5$ cm. Therefore, the buffer
region $V_1'$ lies between $x=7.5$ cm and $x=10$ cm. While $V_1$ is pre-determined at the start of
the simulation and fixed for all energy groups, $V_1'$ does not necessarily coincide for all
energy groups and coordinate axes.

@@

All source code for the hybrid $S_N$-diffusion model is hosted on my Moltres GitHub fork
repository\footnote{\url{https://github.com/smpark7/moltres/tree/sn-method}} as of the time of
writing. This branch may be merged into the original upstream Moltres GitHub repository in the near
future.

@@

The \gls{SAAF} $S_N$ method defines $G$ energy groups and $N_d$ discrete directions for a total
of $G\times N_d$ angular flux variables $\Psi_{g,d}$. The Moltres $S_N$ model groups the
$\Psi_{g,d}$ variables by energy group into $G$ array variables. An array variable is a
set of standard field variables to help with dealing with setting up simulations with high
dimensionality. Each standard variable of an array variable is referred to as a component in the
array variable. In this work, I opted for
2nd-order Lagrange shape functions to approximate the $\Psi_{g,d}$ variables in the finite-element
analysis. The 2nd-order approximation showed better convergence rates in mesh refinement studies.

@@

The \gls{MOOSE} framework leverages the Eigen \cite{guennebaud_eigen_2010} C++ template library for
linear algebra operations defined in ArrayKernels and ArrayBCs for array variables. The \gls{SAAF}
$S_N$ kernels and BCs contain functions for computing the residual and Jacobian contributions of
each term. The \gls{FEM} numerical solver deems to have reached a converged solution if the sum of
residual contributions falls below a defined tolerance value. All $S_N$ simulations in this work
employ the \gls{PJFNK} method \cite{knoll_jacobian-free_2004} with HYPRE-BoomerAMG
\cite{hypre_hypre_2022} for preconditioning. BoomerAMG is a parallel implementation of the
algebraic multigrid method from the HYPRE linear solver library.

Moltres initializes the drift correction variables $\vec{D}_g$ as auxiliary array variables
with three components corresponding to their $x$, $y$, and $z$ components. Auxiliary variables
encompass all field variables which are not directly determined during the \gls{PJFNK}
calculation stage. The drift and boundary correction variables are computed during the
post-calculation stages using $\Psi_{g,d}$.
I created new kernel classes, \texttt{GroupDrift} and \texttt{VacuumCorrectionBC}, to add the
drift and boundary correction terms to the transport-corrected neutron diffusion solver.

I employed the \texttt{MultiApp} system \cite{gaston_physics-based_2015} to implement the fixed point
iterative coupling between the $S_N$ and neutron diffusion solvers for the hybrid $S_N$-diffusion method.
The \texttt{MultiApp} and \texttt{Transfers} systems from \gls{MOOSE} provide flexible interfaces
for setting up iterative solvers and transferring data between the solvers. A typical hybrid
method simulation on Moltres starts with the neutron diffusion solver for the first estimate of the
neutron flux distribution. Moltres transfers the neutron flux distribution to the $S_N$ solver to
compute the boundary source distribution. The $S_N$ solver then generates its own flux estimate and
computes drift and boundary correction variables for the neutron diffusion solver. Moltres
transfers the correction variables to the neutron diffusion solver to compute its next flux
estimate. This process iterates until the neutron flux distribution converges to within tolerance
values provided in the input file.

@@

The hybrid $S_N$-diffusion method improves on the standard neutron
diffusion method by iteratively applying transport corrections generated from solving the $S_N$
neutron transport
method in subdomains containing highly neutron-absorbing regions (e.g., control rods). By limiting
the size of the subdomain in which $S_N$ subsolver is active, the hybrid method saves on
computational expense at the cost of some accuracy in the regions not treated with transport
corrections. In turn, the neutron diffusion subsolver provides approximate boundary source terms
for the $S_N$ subsolver boundary conditions.

I presented two different formulations of transport correction that I investigated in this work for
implementation in the hybrid method: diffusion correction and drift correction. Diffusion
correction involves replacing the diffusion coefficient in the neutron diffusion equations with
corrected coefficients derived from neutron transport methods. Drift correction involves
introducing additional drift terms to the neutron diffusion equations to provide transport
corrections. I found drift correction to be more promising due to issues with division by zero when
obtaining diffusion corrections.

In my investigations with the hybrid method, I found that the $S_N$ subsolver produces
accurate transport correction values in most of the subdomain. However, the transport correction
values near the $S_N$ subsolver boundary are inaccurate due to the approximate boundary conditions
provided by the diffusion subsolver. Thus, I developed and presented a numerical fix for discarding
inaccurate transport correction values and preserving the smoothness of the neutron flux and flux
gradient solution.

By implementing the hybrid method in Moltres, I leveraged existing neutronics modelling
capabilities and parallel numerical solver libraries available through Moltres and \gls{MOOSE} to
enable full-core, time-dependent reactor simulations on \gls{HPC} clusters.
In the next chapter, I verify the hybrid method against Monte Carlo and $S_N$ neutron transport
methods through various neutronics test cases and discuss some significant observations made in my
investigations.
